---
title: "Punxsutawney Phil is as Accurate as a Coin Flip, but Who's Really on Trial Here?"
author: "Phil Azar"
date: "2019-02-02"
output: html_document
slug: punxsutawney-phil
categories: 
- Media
- Politics
- Bill Murray
- Simple Stat Sunday
tags: []
description: "What is a groundhog to do in an era where clicks matter more than content?"
ReadingTime: 3
---

*Simple Stats Sunday&trade; shares a story around a single data point or statistic. Usually on a Sunday. But this one is on Saturday.*

```{r setup, echo=FALSE,message=FALSE, error=FALSE}
library(ggplot2)
library(readr)
library(png)
library(stringr)
library(grid)

y <- c(-50:120)
x <-  c(-50:120)
axis <- expand.grid('x'=x, 'y'=y)
image <- png::readPNG(source='/Users/acazar/Desktop/blog/projects/groundhogs_day/data/images/groundhog_02.png')
pic <- rasterGrob(image, interpolate = TRUE)

```

This morning Punxsutawney Phil awoke with the triumphant forecast of an Early Spring. His decision is brief and final, and welcomed news to many in the Midwest still thawing out from last week's polar vortex. Since 1887, Punxsutawney Phil has predicted the weather into two  categories - Early Spring or Long Winter. But in an age of increasing reductionism in our societal discourse, is this not another attempt to distill the complex into a watered-down bullet point to be added among the day's media feed without proper forum for rationale and debate?  


```{r accuracy, echo=FALSE, message=FALSE, error=FALSE}
ggplot(axis, aes(x=x, y=y)) + 
  geom_hline(yintercept = 50) +
  geom_vline(xintercept = 50) +
  annotate("rect", xmin=50,xmax=100, ymin=50, ymax=100, fill = '#32CD32', alpha =.5) +
  annotate("rect",xmin=0,xmax=50, ymin=0, ymax=50, fill = '#32CD32', alpha = .5) +
  annotate("rect",xmin=0,xmax=50, ymin=50, ymax=100, fill = 'firebrick1', alpha = .5) +
  annotate("rect",xmin=50,xmax=100, ymin=0, ymax=50, fill = 'firebrick1', alpha = .5) +
  annotate("text", x=25, y=75, color="white", fontface='bold', label = "36.6%", size =8) +
  annotate("text", x=75, y=75, color="white", fontface='bold', label = "36.6%", size =8) +
  annotate("text", x=25, y=25, color="white", fontface='bold', label = "14.6%", size =8) +
  annotate("text", x=75, y=25, color="white", fontface='bold', label = "12.2%", size =8) +
  ggtitle("What Actually Happened") +
  ylab("What Phil Predicted") +
  annotate("text", x=-5, y=75, color="grey30", fontface='bold', label = str_wrap("Long Winter",4), size =3) +
  annotate("text", x=-5, y=25, color="grey30", fontface='bold', label = str_wrap("Early Spring",4), size =3) +
  annotate("text", x=25, y=105, color="grey30", fontface='bold', label = str_wrap("Early Spring",4), size =3) +
  annotate("text", x=75, y=105, color="grey30", fontface='bold', label = str_wrap("Long Winter",4), size =3) +
   scale_y_continuous(limits=c(0,110)) +
  scale_x_continuous(limits=c(-10,100)) +
  annotation_custom(pic, xmin=44, xmax=57, ymin=38, ymax=62) +
  theme_minimal() +
  labs(caption = "Based on Average February Temperatures in Boston (1978-2018).") +
  theme(axis.title.y = element_text(face = "bold", color='grey30', size =14), 
        axis.title.x = element_blank(), 
        axis.text = element_blank(), 
        panel.grid = element_blank(), 
        plot.caption = element_text(size =5),  
        plot.title = element_text(hjust=.5, face="bold", color='grey30', size =14)) 
```

By taking the average Feburary temperatures in Boston, MA, since 1978 and categorizing years falling above/below the rolling average temperature into Early Spring/Long Winter, Punxsutawney Phil is accurate about 51% of the time. Or, close to a coin flip. However, measuring Phil with the same simplistic yardstick he was given (and didn't ask for) will not progress the greater conversation we need to have.

```{r forecast, echo=FALSE, message=FALSE, error=FALSE}
ggplot(axis, aes(x=x, y=y)) +
  geom_ribbon(data = subset(axis, x <= 100), aes(ymin = (75/40)*x -112.5, 
                                                ymax = (45/40)*x-37.5,x=x), fill='grey', alpha=.5 ) +
  scale_y_continuous(limits=c(0,100)) +
  scale_x_continuous(limits=c(0,120)) +
  geom_vline(xintercept = 55, size=2, color='grey', alpha=.8) +
  geom_hline(yintercept = 0, size =.5, color='grey', alpha=.8) +
  annotation_custom(pic, xmin=42, xmax=75, ymin=-10, ymax=40) +
  theme_minimal() +
  annotate("text", x =20, y=100, label = "Early Spring", color = 'dodgerblue', 
                      size = 7, fontface='bold', family='Helvetica') +
  annotate("text", x =90, y=100, label = "Long Winter", color = 'black', 
           size = 7, fontface='bold', family='Helvetica') +
  annotate("text", x =105, y=77, label = "79%", color = 'black', 
           size = 5, fontface='bold', family='Helvetica') +
  labs(caption='Punxsutawney Phil Predictions 1887-2019') +
  theme(axis.title.x = element_blank(), 
        axis.title.y = element_blank(), 
        axis.text = element_blank(), 
        panel.grid = element_blank(), 
        plot.caption = element_text(size =5)) 
```

Of course Punxsutawney Phil predicts a Long Winter 79% of the time. His pessimistic outlook is molded by his experience with a media cycle that demands quick takes without first dissecting the facts. Phil struggles with the demand for perfection in a field he has no agency to affect positive change. Human caused global warming makes forecasting the weather more difficult each year and, yet, every February 2nd the world casts their eyes towards Phil to ignore the grim in favor of the benign. The prediction of Long Winter - one of only two choices given to him - is both a protest and wake up call. Double P exchanges accuracy for the realization of our bleak future if we continue on our current trajectory. 

This morning's assurance of an Early Spring can only be understood from the perspective of Punxsutawney Phil. Unfortunately, Phil's predictions are fed to us through the bureaucratic Punxsutawney Groundhog Club Inner Circle, a select group of top hat adorning power players with undisclosed interests. The corporate filter placed upon Phil transforms his predictions into bite-sized entertainment. What is a groundhog to do in an era where clicks matter more than content? Burrow onward and upward. Punxsutawney Phil will return next year and continue his quest to speak truth in the face of power. The only question is will we be listening? 









