---
title: "Onside Kicks Are More Hopeless Than Hopeful"
author: "Phil Azar"
date: "2019-01-20"
slug: onside-kicks
categories: 
- Football
- Sports
- Simple Stat Sunday
tags: []
description: "The game ain't played how it use to. Especially for kickers."
ReadingTime: 2
---

*Simple Stats Sunday&trade; shares a story around a single data point or statistic. Usually on a Sunday.*

```{r data, echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)

comebacks <- suppressMessages(readr::read_csv('/Users/acazar/Desktop/blog/projects/nfl_onside_kicks/data/comebacks_2010_18.csv'))
years = c( "2010", "2011", "2012", "2013", "2014", "2015", "2016", "2017", "2018")

comebacks$Year <- factor(comebacks$Year,levels = c( "2010", "2011", "2012", "2013", "2014", "2015", "2016", "2017", "2018")
 )
osk = c(53,57,60, 62,58, 66,61,56,52)
osr = c(9,10,7,11,8,9,7,12,4)

onside_df = data.frame('years' = years,
                       'osk' = osk, 
                       'osr'= osr)

onside_df$success_r = onside_df$osr/onside_df$osk


```

Last Sunday, the Los Angeles Chargers finally scored against the New England Patriots with 7:40 to go in the fourth quarter when Philip Rivers found tight end Vigil Green alone in the end zone. After a succesful two-point conversion, the LA franchise found themselves down by 19 against Bill Bellicheck, Tom Brady, and the clock. Fans between San Diego and Orange County were doing the mental calculations to add up to a Charger's victory. Albeit already improbable, a Charger's win depended on recovering multiple onside kicks, which made their comeback even less likely.

```{r success_plot, echo=FALSE}

plot <- ggplot(onside_df, aes(x=years, y=success_r, group=1, label=scales::percent(success_r))) +
        geom_line(size=2, color='dodgerblue') +
        geom_text(aes(label = stringr::str_wrap('1 in 5 onside kicks were recovered in 2017', 17), x='2017', y=.25),
                  color='black', 
                  size=3, 
                  fontface='bold',
                  family='Helvetica') +
        scale_y_continuous(labels = scales::percent, limits=c(0,.3)) +
        ggtitle('Onside Kicks Hit Rock Bottom in 2018') +
        ylab('Onside Kick Recovery Rate by Kicking Team') +
        xlab('Season') +
        geom_text(aes(x='2018',y=.18, label = stringr::str_wrap('NFL Rule Change', 5)), 
                  family='Helvetica', 
                  colour= 'black', 
                  fontface='bold',
                  size=3) +
        labs(caption= 'Source: Pro Football Reference') +
        theme(plot.title = element_text(hjust=.5, family = 'Helvetica', face='bold'), 
              axis.title.x = element_text(hjust=.5, family='Helvetica', face='bold'), 
              axis.title.y = element_text(hjust=.5, family='Helvetica', face='bold'),
              axis.text = element_text(family='Helvetica', face='bold'),
              panel.grid.minor = element_blank(), 
              panel.background = element_rect(fill='grey80'), 
              strip.background = element_rect(fill='grey80'),
              plot.background = element_rect(fill='grey80'),
              panel.border = element_blank(), 
              plot.caption = element_text(family='Helvetica', size=5))

plot


```


In 2018, only 7.7% of onside kicks were recovered by the kicking team - the lowest success rate ever. Between 2010 and 2017, the average recovery rate of an onside kick was 16%. In 2017 - the Golden Age of Onside Kicking (probably) - the recovery rate peaked to over 21%. The decline in teams' ability to maintain possession after a score is not due to poor kicking or bad hands, but to a simple rule change. 

The NFL kickoff is considered the most dangerous play in the sport. Last offseason NFL owners considered many rule changes to the play; even doing away with kickoffs altogther. One rule they agreed upon disallowed the kicking team from getting a 5-yard running start on onside kicks. The running start gave the kicking team momentum into the flat footed recieving team 10 yards away. More momentum meant faster speed to recover the ball. But it also meant more collisions and injuries. Now, the kicking team must line up within 1 yard of the kickoff, and kick it the minimum 10 yards where their opponent waits patiently. Kicking teams inability to build up speed to get down the field fast enough has lead to an onside kick extinction. After the first season of the rule, however, it does not appear to be that all decisive.


```{r 4QW, echo=FALSE, message=FALSE, warning=FALSE}

plot_2 <- comebacks %>% 
  group_by(Year) %>% 
  summarise(comebacks = sum(`4QC`)) %>% 
ggplot(., aes(x=Year, y=comebacks, group=1, label=comebacks)) +
        geom_col(fill='dodgerblue', width=.5) +
        geom_text(aes(label=comebacks), 
                  nudge_y = 3, 
                  color='black',
                  size=3, 
                  fontfamily='bold',
                  family='Helvetica') +
scale_y_continuous(limits=c(0,80)) +
ggtitle('4th Quarter Comebacks Still Likely to Happen') +
  xlab('Season') +
  ylab('4th Quarter Comebacks') +
   labs(caption= 'Source: Pro Football Reference') +
        theme(plot.title = element_text(hjust=.5, family = 'Helvetica', face='bold'), 
              axis.title.x = element_text(hjust=.5, family='Helvetica', face='bold'), 
              axis.title.y = element_text(hjust=.5, family='Helvetica', face='bold'),
              axis.text = element_text(family='Helvetica', face='bold'),
              panel.grid.minor = element_blank(), 
              panel.background = element_rect(fill='grey80'), 
              strip.background = element_rect(fill='grey80'),
              plot.background = element_rect(fill='grey80'),
              panel.border = element_blank(), 
              plot.caption = element_text(family='Helvetica', size=5), 
              panel.grid.major.x = element_blank())

plot_2

```



At the beginning of the 2018 NFL season [many bemoaned the rule change](https://www.usatoday.com/story/sports/ftw/2018/09/10/the-nfls-new-kickoff-rules-have-made-onside-kicks-impossible/111299536/), but at the end of the season, the dearth of onside kick recoveries did not decide winners and losers. In 2018, there were 71 comeback wins, which is slightly above the 65 average since 2010. Onside kicks are more acts of desperation than a winning strategy. No coach hopes to be down by a score without possession and time on the clock. At least no coach who is employed. 

Fans, on the other hand, cling to the bit of hope that echoes in living rooms across the country *'then we recover the onside kick and...'*. During today's AFC and NFC championship games, if you find yourself saying this like many Charger fans last week,  
the game and the ball might already be out of your hands.


*Data and Statistics from [Pro Football Reference](https://www.pro-football-reference.com)*









