---
title: Everyone Hates Rand
author: Phil Azar
date: '2018-04-15'
slug: everyone-hates-rand
categories: []
tags: ["networks", "politics", "twitter", "network analysis", "Senate"]
description: "Analyzing the followers of all 100 U.S. Senators shows a network of bipartisanship, cliques and a fuzzy haired libertarian"
---
```{r, include=FALSE,echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(ggnetwork)
library(stringr)
library(ggiraph)
library(statnet)

senate_gg <- readRDS('/Users/acazar/Desktop/blog/projects/congress_twitter/data/senate_gg.rds')
network_plot <- ggplot(senate_gg, aes(x = x, y = y, xend=xend, yend=yend, color = party,
                              tooltip=vertex.names)) +
  geom_edges(aes(linetype = line), curvature=.4, alpha=.3, size=.17, show.legend = FALSE) +
  geom_nodes(alpha= .4, aes(fill =party, size =betweenness)) +
  scale_color_manual(values=c('Democrat'='dodgerblue', 'Republican'='red', 'Independent'='forestgreen', 'NA'='grey')) +
  theme_blank() +
  scale_linetype_manual(values = c('solid'='solid', 'dotted'='dotted')) +
  geom_nodetext(aes(label=vertex.names, x=x,y=y), fontface='bold', size =3, nudge_y=.025) +
  guides(color="none", fill='none') +
  theme(legend.position = 'bottom') +
  geom_text(x=.55, y=.12, size= 2.25, color='grey50', alpha= .8,
            label='dotted lines show connections between different parties') +
  geom_text(x=.40, y =.075, size =2.25,alpha=.8, color ='grey50', label='solid lines show connections to same party')




```

     Witty jokes, Russian operatives, breaking news and Jim Carrey. Twitter is a thought jacuzzi with a welcoming jet for every one. The timeline offers a warm place to scroll through nodding heads reminding us how right we are. Since the election, my feed of jokes and MLB trade rumors rotted into political discourse. Each morning I took in a fresh political outrage with my coffee. It was exhausting. I began unfollowing accounts. Too hyperbolic? Gone. Doesn't agree with me? Next. Retweets 2010 Sam Bradford highlights? You stay. 
     
     The selection of your echo chamber reminds us of what we want to hear as much as it is portraying to the world the a version of ourselves we like best. No one uses social media more for this purpose than politicians, who transform twitter into a stage to shout to their voter base. Who they follow - and not follow - reveals how they want to be perceived. Analyzing the follower network of all 100 U.S. Senators shows a network of bipartisanship, cliques and a fuzzy haired libertarian. 
     
     
```{r, echo=TRUE, warning=FALSE}
ggiraph(code={print(network_plot)})
```
     
    
     
     
     
     
     
     
     